---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

The variables in this dateset are:
Class: whether the person was 1st class, 2nd class, 3rd class, or a crew member on the titanic
Sex of the person
Age: whethere they were a child or an adult
Survived: if they survived the titanic or not
n: the number of people in that specific group

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
- Are there any differences?
  - (Your response here)
- If yes, what might account for those differences?
  
  The wikipedia article lists that a total of 2224 people were on-board the titanic. 1517 would die in the ship sinking, while 704 people survived. The date set we are given lists a total of 2201 people, which is fairly close to wikipedia estimate of total people. There are a few possible reasons for this discrepancy. The first being that perhaps certain records may have been lost over time. Another reason could be that their families requested that this information be kept anonymous. Since, the difference between the size of this data set and number of officially recorded passengers we can probably be fairly confident in the data.
  
  - ...

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>% 
  ggplot()+
  geom_col(mapping = aes(x = Class, y = n, fill = Sex))

```

**Observations**:

From looking at this graph, it is shown that among the passengers more females survived than males. However, among the crew the survivors were almost entirely male. It's hard to know if this was due to the crew was predominately male or because the male crew were more eager to secure spots in the lifeboats, since the graph doesn't show the breakdown of people that didn't survive. The same is true of the passengers. It is hard to analyze the data, since we don't know the percentages of males and females in each group that survived.

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop

```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot()+
    geom_col(mapping = aes(x = Class, y = Prop, fill = Sex))
```

**Observations**:

- Write your observations here.
- Is there anything *fishy* going on in your plot?

From looking at this chart, it looks like females did tend to survive at higher rates than males. The largest proportion of survivors were from first class, followed by second, and then third. Crew members had the lowest survival rate, especially male crew members.

The "fishy" part of this plot if that the proportions are greater than 1. This is due to the fact that in the calculation the proportions are getting lumped together.

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col() +
  facet_grid(~ Age)
```

**Observations**:

From looking at this data, it we can see that among adults females survived at much higher rates than males. This difference is most striking when looking at the proportion of males that survived in 2nd class. Among children however, the breakdown is fairly even. Overall, children survived at much higher rates than adults, especially if they were in first or second class.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.

```{r}

```

