---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
help(gapminder)
glimpse(gapminder)
```

**Observations**:

The variables in this data set are:
country, continent, year (1952 - 2007 in 5 year increments), life expectancy (in years), population, GDP per capita (USD, adjusted for inflation)

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
years_list <-
  gapminder %>% 
  group_by(year) %>% 
  pull(year)

year_max <-
  max(years_list)
year_min <-
  min(years_list)
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot +
  geom_boxplot(mapping = aes(x = continent, y = gdpPercap)) +
  scale_y_log10()
```

**Observations**:

It's difficult to read this chart, because the scaling on the graph is messed up by the outlier with an over 90000 gdp per capita. However, I can still see that Oceania has the highest overall gdp, whole Africa and Asia are on the lower end. There are also many outliers represented on the graph as points. Africa has a few, Americas have 3, Asia has 2, and Europe has 1.

**Difficulties & Approaches**:

It's difficult to read this chart, because the scaling on the graph is messed up by the outlier with an over 90000 gdp per capita. I attempted to look through the reading on box plots in order to figure out how to fix the scaling, however I wasn't able to figure it out.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
gapminder %>% 
  filter(year == year_min) %>% 
  group_by(continent) %>% 
  filter(gdpPercap >= quantile(gdpPercap, 0.75) + 1.5*IQR(gdpPercap))
```

**Observations**:

From looking at the list of outliers, I can see that there are a few countries that I would expect on the list, for example the US, Switzerland, and South Africa. There were some smaller countries that I was suprised to see on the list and didn't know anything about. For example, Bahrain and Kuwait. But through some quick online digging, it looks like those countries economies are largely driven by oil reserves which could be an explanation for their high gdp per capitas.


*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
gapminder %>% 
  filter(year %in% c(year_min,year_max)) %>% 
  filter(gdpPercap < 30000) %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = continent, y = gdpPercap)) +
  facet_wrap(~ year) +
  geom_point(
    data = . %>% filter(country %in% c("United States", "Angola", "Bahrain", "Canada", "Gabon", "South Africa", "Switzerland","Venezuela")),
    mapping = aes(x = continent, y = gdpPercap, color = country),
    size = 2
  ) +
  scale_y_log10()

```

**Observations**:

From looking at the most recent data, it looks like there are fewer outliers for most of the continents. It looks like the americas have 2 outliers, while Africa has around 5 or 6. Asia and Europe have the largest IQRs, which means that the gdp's for countries in those continents are much more varied than the countries in Oceania. 

Compairing between 

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1} 
gapminder %>% 
  filter(year == year_max) %>% 
  ggplot +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp)) +
  scale_y_log10()
```

The reason I made this graph was to understand the relationship between gdp per capita and life expectancy. I wanted to understand the extent to which wealth could improve life expectancy.From looking at the graph, it looks like for very low gdp's (below 5000) life expectancy is quite low. However, after gdp increases past 15000 the relationship gets weaker. 


```{r q5-task2}
gapminder %>% 
  filter(country %in% c("Kuwait","United States")) %>% 
  ggplot +
  geom_point(mapping = aes(x = year, y = gdpPercap, color = country))
```

I made this graph to look at how countries who's economies rely on natural resources, such as oil reserves, fare over time. Since there are a finite amount of easily reachable oil reserves it's interesting to look at the gdp of such a country over time. For Kuwait, it looks like they started off with a gdp per capita over 4 times higher than the US in 1952, and by 1980 it fell to a gdp per capita similar to that of the US. It would be interesting to look into whether the country still primarily exports oil, or if it has greatly diversified it's economy.

```{r q5-task3}
gapminder %>% 
  filter(country %in% c("France", "United States", "China","Australia")) %>% 
  ggplot +
  geom_point(mapping = aes(x = year, y = lifeExp, color = country)) + 
  scale_y_log10()
```

The reason I made this graph to examine the effect on life expectancy that the great leap forward had in China. I had previously learned in history class that it was one of the most devastaing famines in history, and as shown in the graph there is a noticable dip in life expentancy in the years that followed the great leap forward. 
