---
title: "SAT and College Grades"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: How do we apply hypothesis testing to investigating data? In
this challenge you'll practice using hypothesis testing tools to make
sense of a dataset.

*Reading*: - [Harvard Study Says SATs Should Be Optional: Here's
Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why)
(Optional); easy-to-read news article on colleges going SAT-free -
[Norm-Referenced Tests and Race-Blind
Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university)
(Optional); technical report on relationship between the SAT/ACT and
non-academic factors

*Credit*: This is based on a [case
study](http://onlinestatbook.com/2/case_studies/sat.html) originally
prepared by Emily Zitek, with data collected through the research of
Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
```

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

*Background*: Every year about 2 million students take the Scholastic
Aptitude Test (SAT). The exam is
[controversial](http://www.nea.org/home/73288.htm) but [extremely
consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html).
There are many claims about the SAT, but we're going to look at just
one: Is the SAT predictive of scholastic performance in college? It
turns out this is a fairly complicated question to assess---we'll get an
introduction to some of the complexities.

# Obtain the Data

<!-- -------------------------------------------------- -->

### **q1** Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
filename_sat <- "./data/df_sat.xls"

df_sat <-
  read_xls(filename_sat)

## TODO: Do your "first checks"
glimpse(df_sat)
```

**Observations**:

-   Fill in the following "data dictionary"

| Column     | Meaning                |
|------------|------------------------|
| `high_GPA` | high school GPA        |
| `math_SAT` | SAT Math score         |
| `verb_SAT` | SAT Verbal score       |
| `comp_GPA` | computer science GPA   |
| `univ_GPA` | overall university GPA |

-   What information do we have about these students?
    -   We have information about their high school and university GPAs
        as well as test scores
-   What kinds of information *do we not have* about these students?
    -   We do not have information about background (ethnicity, income)
        or about qualitative college admissions factors (extracurricular
        activities, leadership experience etc.)
-   Based on these missing variables, what possible effects could be
    present in the data that we would have *no way of detecting*?
    -   All of the missing variables could potentially affect the
        outcome of the data

# Analysis with Hypothesis Testing

<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the
first based on hypothesis testing of correlation coefficients, and the
second based on fitting a regression model and interpreting the
regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)

df_composite
```

## View 1: Correlations

<!-- ----------------------------------------------------------------------- -->

### **q2** Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`. "

```{r q2-task}
df_pivot <-
  df_composite %>%
  pivot_longer(
    names_to = c("score_type"),
    values_to = "value",
    cols = c(- univ_GPA)
  )

df_pivot

df_pivot %>%
  filter(score_type %in% c("high_GPA", "both_SAT")) %>% 
  ggplot() +
  geom_point(mapping = aes(x = value, y = univ_GPA)) +
  facet_wrap(~score_type, scales ="free")
```

**Observations**:

-   What relationship do `univ_GPA` and `both_SAT` exhibit?
    -   It seems to exhibit a moderately strong positive correlation
-   What relationship do `univ_GPA` and `high_GPA` exhibit?
    -   It seems to exhibit a moderately strong positive correlation.
        The correlation appears to be stronger than the first graph.

### Hypothesis Testing with a Correlation Coefficient

<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation
coefficient. The idea is to set our null hypothesis to the case where
there is no correlation, and test to see if the data contradict that
perspective. Formally, the null (H0) and alternative (HA) hypotheses
relating to a correlation coefficient between two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the
assumption that `X, Y` are both normally distributed. First, let's check
to see if this assumption looks reasonable for our data.

### **q3** Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}

df_composite %>% 
  ggplot(aes(both_SAT)) +
  geom_histogram(bins = 6)

df_composite %>% 
  ggplot(aes(high_GPA)) +
  geom_histogram(bins = 10)

df_composite %>% 
  ggplot(aes(univ_GPA)) +
  geom_histogram(bins = 10)

df_composite
```

**Observations**:

-   To what extent does `both_SAT` look like a normal distribution?
    -   It does not look like a normal distribution
-   To what extent does `high_GPA` look like a normal distribution?
    -   It does not look like a normal distribution
-   To what extent does `univ_GPA` look like a normal distribution?
    -   It does not look like a normal distribution

Keep in mind your findings as you complete q4.

### **q4** Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA

high_GPA <-
  df_composite %>% 
  pull(high_GPA)

univ_GPA <-
  df_composite %>% 
  pull(univ_GPA)

both_SAT <-
  df_composite %>% 
  pull(both_SAT)

cor.test(high_GPA, univ_GPA)
cor.test(both_SAT, univ_GPA)

```

**Observations**:

-   Which correlations are significantly nonzero?
    -   both correlations are significantly nonzero
-   Which of `high_GPA` and `both_SAT` seems to be more strongly
    correlated with `univ_GPA`?
    -   high_GPA appears to be more strongly correlated with univ_GPA
-   How do the results here compare with the visual you created in q2?
    -   It matches the observation that visually high_GPA seems to have
        a stronger correlation with univ_GPA than both_SAT
-   Based on these results, what can we say about the predictive
    capabilities of both `high_GPA` and `both_SAT` to predict
    `univ_GPA`?
    -   Based on this, high_GPA might be a better predictor than
        both_SAT

Finally, let's use the bootstrap to perform the same test using
*different* assumptions.

### **q5** Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
## TODO: Use the bootstrap to compute a confidence interval for corr[high_GPA, univ_GPA]
n_bootstrap <- 1000

bootstrap_corr <- numeric(n_bootstrap)
for (i in 1:n_bootstrap) {
  sample_idx <- sample(length(high_GPA), replace = TRUE)
  sample_high_GPA <- high_GPA[sample_idx]
  sample_univ_GPA <- univ_GPA[sample_idx]

  bootstrap_corr[i] <- cor(sample_high_GPA, sample_univ_GPA)
}

# Compute confidence interval
lower_ci <- quantile(bootstrap_corr, 0.025)
estimate <- quantile(bootstrap_corr, 0.5)
upper_ci <- quantile(bootstrap_corr, 0.975)

tibble(
  lower_ci = lower_ci,
  estimate = estimate,
  upper_ci = upper_ci)
```

**Observations**:

-   How does your estimate from q5 compare with your estimate from q4?
    -   The estimate is very similar to the estimate from q4
-   How does your CI from q5 compare with your CI from q4?
    -   The CI is very similar to that of q4

*Aside*: When you use two different approximations to compute the same
quantity and get similar results, that's an *encouraging sign*. Such an
outcome lends a bit more credibility to the results.

## View 2: Modeling

<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study
the relationship among more variables we can instead use a fitted model.
Using a model, we can also help assess whether it is *worthwhile* to
measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model

<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a
model, we can express our hypotheses in terms of the model parameters.
For instance, if we were interested in whether $X$ has an affect on $Y$,
we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on
$Y$. Let's apply this idea to relating the variables `univ_GPA` and
`high_GPA`. Luckily R has built-in tools to construct a confidence
interval on the $\beta$'s in a regression [1]; we'll simply use those
tools rather than do it by hand.

### **q6** Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <-
  lm(
    data = df_train,
    formula = univ_GPA ~ both_SAT
  )

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

tibble(
  error_Train = mse(fit_basic, df_train),
  error_Validate = mse(fit_basic, df_validate))
```

**Observations**:

-   What is the confidence interval on the coefficient of `both_SAT`? Is
    this coefficient significantly different from zero?
    -   the confidence interval does not include zero, which is good.
        The lower end of the confidence interval for the coefficient is
        very close to zero however.
-   By itself, how well does `both_SAT` predict `univ_GPA`?
    -   From looking at the mse, it looks like the error is relatively
        small compared to the GPA scale. However, from computing
        correlations above perhaps there is a better predictor that
        could be incorporated into the model as well.

Remember from `e-model03-interp-warnings` that there are challenges with
interpreting regression coefficients! Let's investigate that idea
further.

### **q7** Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone

fit_both <-
  lm(
    data = df_train,
    formula = univ_GPA ~ both_SAT + high_GPA
  )
## NOTE: The following computes confidence intervals on regression coefficients
fit_both %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

tibble(
  error_Train = mse(fit_both, df_train),
  error_Validate = mse(fit_both, df_validate))
```

**Observations**:

-   How well do these models perform, compared to the one you built in
    q6?
    -   Compared to the previous model, this new model that adds in high
        school GPA to the formula performs significantly better (almost
        30%)
-   What is the confidence interval on the coefficient of `both_SAT`
    when including `high_GPA` as a predictor?? Is this coefficient
    significantly different from zero?
    -   the confidence interval for the coefficient of both_SAT includes
        0 in the interval, so it is not significantly greater than 0.
-   How do the hypothesis test results compare with the results in q6?
    -   (Your response here)

## Synthesize

<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### **q8** Using the results from all previous q's, answer the following questions.

**Observations**:

-   Between `both_SAT` and `high_GPA`, which single variable would you
    choose to predict `univ_GPA`? Why?
    -   Out of the two, I would choose high_GPA because the fit from
        high_GPA has a lower error and also it has a higher correlation
        with univ_GPA.
-   Is `both_SAT` an effective predictor of `univ_GPA`? What specific
    pieces of evidence do you have in favor of `both_SAT` being
    effective? What specific pieces of evidence do you have against?
    -   adding both_SAT on to high_GPA does create a more accurate fit.
        From looking a the mse value for both_SAT alone, it is
        relatively small on the scale of GPA. However it doesn't perform
        as well as other values such as high_GPA.

# End Notes

<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more
information see this [Wiki
article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
